The Key to a Data Parallel Compiler

### 2015-11-11 2016-pldi/2016-pldi-key_to_data_parallel_compiler.tex
Reorganize things in the PLDI paper to match the new outline.
Add in a table for compiler passes.

\begin{itemize}[noitemsep]
\item We describe a method of computing over arbitrary sub-trees selected by their parent-child 
relationships in a data-parallel manner using the Key operator and Node Coordinates.
\item We situate this technique into a broader, language-driven compiler construction strategy 
that enables the development of parallel by construction compilers that exhibit desirable programmability 
features in terms of code size, complexity, portability, and maintainability. 
\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and 
report on the results.
\item We compare these techniques to the Nanopass approach to compiler design in terms of programmer 
productivity and compiler runtime performance. 
\item We enumerate a set of programming techniques, patterns, and strategies that complement the 
Key operator and Node Coordinates in the construction of tree transformations in general and give some 
intuitions on the use of these techniques on arbitrary transformations.
\end{itemize}
%[cf]
%[of]:Introduction
\section{Introduction (Old)}

\begin{itemize}[noitemsep]
\item We introduce node coordinates as a means to locally compute inter-node properties without using complex 
control flow.
\item We demonstrate methods to perform arbitrary computation over sub-trees partitioned by these 
inter-node properties in a pure data-flow, data-parallel style.
\item The presented methods are direct and concise; the data structures, simple.
\item These methods require no specialized operations, but instead, use only established, general-purpose 
array programming.
\end{itemize}
%[cf]
%[cf]
%[of]:A Definition of Node Coordinates
\section{A Definition of Node Coordinates}

To understand how we construct these coordinates we must consider how we represent the AST.
%[cf]
%[of]:Encoding the AST
\section{Encoding the AST}

@@

Note especially that all inter-node information lives in the depth vector, but that this information requires 
non-local access as described in the previous section to use it. Constructing node coordinates fixes this issue.
%[cf]
%[of]:Constructing Node Coordinates
\section{Constructing Node Coordinates}

@@

A careful study of the definition of a node coordinate and the above construction should 
reveal why this works. Intuitively, we are creating a multi-dimensional space or a number 
system in which each digit place or dimension contains or circumscribes a smaller space in 
which are contained all the descendant nodes that appear lower in the tree. Each coordinate 
is a sort of special path through the tree encoded to have desirable properties relative to 
other paths.
%[cf]
%[of]:Operations on Node Coordinates
\section{Operations on Node Coordinates}

@@

We will use these keys to perform computation over the AST and accomplish our tasks of function lifting and 
expression flattening in the next section, but we make a final note here, that we can imagine many other 
operations which might be used throughout the compiler when we care about how nodes relate to each 
other. The use of operators like reduction, scan, inner and outer products allow us to obtain the necessary 
information from the node coordinate matrix. Sometimes these results might not be obvious, but the 
example of the “belongs to” relationship given above demonstrates a pattern that arises repeatedly 
throughout our data parallel compiler, and is therefore particularly useful to understand. 
%[cf]
%[of]:Computing with Node Coordinates
\section{Computing with Node Coordinates}

Given the above keys and grouping information, we now have all the information we need to do 
function lifting and expression flattening, but we lack a suitable operator to complete the task. 
Indeed, it isn’t immediately apparent that a single, general purpose operator over arrays exists 
which can accomplish such a task. However, the Key operator, while masquerading as a useful business 
analytics operator, represents just such a useful tree transformation tool.

In our case, we use either Fk or Ek as our keys applied to the corresponding AST. We also will drop off 
the first row in each AST using 1↓ since this node “contains” everything. In a complete AST this is 
usually the Module boundary node which contains the entire set of functions and values in the module.
%[cf]
%[of]:Function Lifting
\subsection{Function Lifting}

@@

The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we 
are assuming that the order of evaluation is right to left, which means that the above order is actually 
backwards of the desired order. During recombination, we simply reverse these orders and this fixes 
that problem. More work would be required to take into consideration a specific precedence hierarchy. 
%[cf]
%[cf]
%[of]:Scaling up, Considerations and Thoughts
\section{Scaling up, Considerations and Thoughts}

Finally, it is interesting to note that the simplified control flow gives rise to some positive side-effects. 
Besides brevity and directness, the complexity of each operation and the composition of operations is 
well known. We have not yet tried to do this, but it is straightforward to derive a worst-case complexity 
bound on code written in this style, so simple, in fact, that it can be accomplished automatically. It is 
conceivable then, that the compiler could produce not only an executable program from code written 
in this style, but also a complexity bound on its execution. 
%[cf]
%[of]:Related Work
\section{Related Work}

@@

>>>>

\begin{itemize}[noitemsep]
\item We describe a method of computing over arbitrary sub-trees selected by their parent-child 
relationships in a data-parallel manner using the Key operator and Node Coordinates.
\item We situate this technique into a broader, language-driven compiler construction strategy 
that enables the development of parallel by construction compilers that exhibit desirable programmability 
features in terms of code size, complexity, portability, and maintainability. 
\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and 
report on the results.
\item We compare these techniques to the Nanopass approach to compiler design in terms of programmer 
productivity and compiler runtime performance. 
\item We enumerate a set of programming techniques, patterns, and strategies that complement the 
Key operator and Node Coordinates in the construction of tree transformations in general and give some 
intuitions on the use of these techniques on arbitrary transformations.
\end{itemize}
%[cf]
%[of]:Generalized Sub-tree Computation
\section{Data Parallel Sub-tree Computation}

In our case, we use either Fk or Ek as our keys applied to the corresponding AST. We also will drop off 
the first row in each AST using 1↓ since this node “contains” everything. In a complete AST this is 
usually the Module boundary node which contains the entire set of functions and values in the module.
%[cf]
%[of]:Encoding the AST
\subsection{Encoding the AST}

@@

Note especially that all inter-node information lives in the depth vector, but that this information requires 
non-local access as described in the previous section to use it. Constructing node coordinates fixes this issue.
%[cf]
%[of]:Node Coordinates
\subsection{Node Coordinates}

%[of]:A Definition of Node Coordinates
\subsubsection{A Definition of Node Coordinates}

To understand how we construct these coordinates we must consider how we represent the AST.
%[cf]
%[of]:Constructing Node Coordinates
\subsubsection{Constructing Node Coordinates}

@@

A careful study of the definition of a node coordinate and the above construction should 
reveal why this works. Intuitively, we are creating a multi-dimensional space or a number 
system in which each digit place or dimension contains or circumscribes a smaller space in 
which are contained all the descendant nodes that appear lower in the tree. Each coordinate 
is a sort of special path through the tree encoded to have desirable properties relative to 
other paths.
%[cf]
%[of]:Operations on Node Coordinates
\subsubsection{Operations on Node Coordinates}

@@

We will use these keys to perform computation over the AST and accomplish our tasks of function lifting and 
expression flattening in the next section, but we make a final note here, that we can imagine many other 
operations which might be used throughout the compiler when we care about how nodes relate to each 
other. The use of operators like reduction, scan, inner and outer products allow us to obtain the necessary 
information from the node coordinate matrix. Sometimes these results might not be obvious, but the 
example of the “belongs to” relationship given above demonstrates a pattern that arises repeatedly 
throughout our data parallel compiler, and is therefore particularly useful to understand. 
%[cf]

%[cf]
%[cf]
%[of]:Case Studies
\section{Case Studies}

%[of]:Table of Compiler Passes
\begin{table*}
\centering
\begin{tabular}{l p{2.4in} l l l}
\textbf{Pass} & \textbf{Description} & \textbf{Core Patterns} 
 & \textbf{Key?} & \textbf{Coordinate?} \\ \hline
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & No & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & No & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & No & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & No & Yes \\
Fold Constants & Converts constant expressions to literals & Amend & No & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & Yes & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No & No \\
Lift Type-checking & Infers some type information at compile time &
 Power Limit/Rank & No & No \\
Allocate Value Slots & Does a form of frame allocation for variables &
 Key & Yes & No \\
Anchor Variables & Resolves lexically scoped variables &
 Key & Yes & No \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & Yes & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
 Key & Yes & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & Yes & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{table:passes}
\end{table*}
%[cf]
%[of]:Function Lifting
\subsection{Function Lifting}

@@

The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we 
are assuming that the order of evaluation is right to left, which means that the above order is actually 
backwards of the desired order. During recombination, we simply reverse these orders and this fixes 
that problem. More work would be required to take into consideration a specific precedence hierarchy. 
%[cf]
%[of]:Lexical Scoping
\subsection{Lexical Scoping}
%[cf]
%[cf]
%[of]:Other Passes
\section{Other Passes}
%[cf]
%[of]:The Co-dfns Compiler
\section{The Co-dfns Compiler}
%[cf]
%[of]:Performance
\section{Performance}
%[cf]
%[of]:Future Work
\section{Future Work}
%[cf]
%[of]:Related Work
\section{Related Work}

@@

%[of]:Unintegrated Sections
\section{Unintegrated Sections}

%[of]:Scaling up, Considerations and Thoughts
\subsection{Scaling up, Considerations and Thoughts}

Finally, it is interesting to note that the simplified control flow gives rise to some positive side-effects. 
Besides brevity and directness, the complexity of each operation and the composition of operations is 
well known. We have not yet tried to do this, but it is straightforward to derive a worst-case complexity 
bound on code written in this style, so simple, in fact, that it can be accomplished automatically. It is 
conceivable then, that the compiler could produce not only an executable program from code written 
in this style, but also a complexity bound on its execution. 
%[cf]
%[of]:Computing with Node Coordinates
\subsection{Computing with Node Coordinates}

Given the above keys and grouping information, we now have all the information we need to do 
function lifting and expression flattening, but we lack a suitable operator to complete the task. 
Indeed, it isn’t immediately apparent that a single, general purpose operator over arrays exists 
which can accomplish such a task. However, the Key operator, while masquerading as a useful business 
analytics operator, represents just such a useful tree transformation tool. 
%[cf]
%[of]:Introduction (Old)
\subsection{Introduction (Old)}

\begin{itemize}[noitemsep]
\item We introduce node coordinates as a means to locally compute inter-node properties without using complex 
control flow.
\item We demonstrate methods to perform arbitrary computation over sub-trees partitioned by these 
inter-node properties in a pure data-flow, data-parallel style.
\item The presented methods are direct and concise; the data structures, simple.
\item These methods require no specialized operations, but instead, use only established, general-purpose 
array programming.
\end{itemize}
%[cf]
%[cf]

%[cf]
